---
title: "Geospatial"
author: "Brian Yandell"
date: "2023-06-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

See <README.md>.

# Day 1: Raster Data

```{r}
HARV_dsmCrop_info <-
  utils::capture.output(
    terra::describe("data/raster/HARV_dsmCrop.tif"))
```

```{r}
(DSM_HARV <- terra::rast("data/raster/HARV_dsmCrop.tif"))
```

```{r}
summary(terra::values(DSM_HARV))
```
## Plot raster data

You must create df from raster data for use in ggplot2,
say using S4 method for `SpatVector` object

```{r}
ggplot2::ggplot(
  as.data.frame(DSM_HARV, xy = TRUE)) +
  ggplot2::aes(x, y, fill = HARV_dsmCrop) +
  # Raster geometry
  ggplot2::geom_raster() +
  # Add Viridis gradient scale
  ggplot2::scale_fill_viridis_c()
```

Quick plot of raster

```{r eval = FALSE}
plot(DSM_HARV)
```

### Look at coordinate reference system

- is it projected or not?
- what units?
- where is this place?

```{r}
terra::crs(DSM_HARV, proj = TRUE)
```

```{r}
terra::minmax(DSM_HARV)
```

Force saving of min and max with object.

```{r}
DSM_HARV <- terra::setMinMax(DSM_HARV)
```

### Plot histogram of data

```{r}
ggplot2::ggplot(
  as.data.frame(DSM_HARV, xy = TRUE)) +
  ggplot2::aes(HARV_dsmCrop) +
  # Histogram
  ggplot2::geom_histogram()
```

```{r}
ggplot2::ggplot(
  as.data.frame(DSM_HARV, xy = TRUE)) +
  ggplot2::aes(HARV_dsmCrop) +
  # Histogram
  ggplot2::geom_density()
```

Plot with breaks. **Caution:** if bins are uneven, it would be helpful to use the
`width` option on `geom_bar()` to show this.

```{r}
custom_bins <- c(300,350,400,450)
```

Using package `dplyr`. Rather than using pipes `%>%` I explicitly call the `dplyr`
routing `mutate` with first argument as the data frame.

```{r}
ggplot2::ggplot(
  dplyr::mutate(
    as.data.frame(DSM_HARV, xy = TRUE),
    fct_elevation = cut(HARV_dsmCrop, custom_bins))) +
  ggplot2::aes(fct_elevation) +
  # Bar chart
  ggplot2::geom_bar()
```

```{r}
ggplot2::ggplot(
  # Cut values into bins based on `custom_bins` defined above.
  dplyr::mutate(
    as.data.frame(DSM_HARV, xy = TRUE),
    fct_elevation = cut(HARV_dsmCrop, custom_bins))) +
  ggplot2::aes(x, y, fill = fct_elevation) +
  # Raster plot
  ggplot2::geom_raster() +
  # Visualize simple feature objects.
  ggplot2::coord_sf() +
  # Add terrain color pallette.
  ggplot2::scale_fill_manual(values = grDevices::terrain.colors(3))
```

Challenge: use 6 bins. Add title and axis labels.

```{r}
custom_bins <- seq(300, 450, length.out = 7)
```

```{r}
ggplot2::ggplot(
  # Cut values into bins based on `custom_bins` defined above.
  dplyr::mutate(
    as.data.frame(DSM_HARV, xy = TRUE),
    fct_elevation = cut(HARV_dsmCrop, custom_bins))) +
  ggplot2::aes(x, y, fill = fct_elevation) +
  # Raster plot
  ggplot2::geom_raster() +
  # Visualize simple feature objects.
  ggplot2::coord_sf() +
  # Add terrain color pallette.
  ggplot2::scale_fill_manual(values = grDevices::terrain.colors(length(custom_bins) - 1),
                             name = "Elevation") +
  # Add title and axes
  ggplot2::ggtitle("NEON Harvard Forest Crop") +
  ggplot2::labs(x = "UTM Easting Coordinates (m)",
                y = "UTM Northing Coordinates (m)")

```

# Day 2: More Raster Data

Load more data

```{r}
DSM_hill_HARV <- terra::rast("data/raster/HARV_DSMhill.tif")
```

Using parameter `alpha` for [transparency level](https://ggplot2.tidyverse.org/reference/scale_alpha.html).

```{r}
ggplot2::ggplot(
  as.data.frame(DSM_hill_HARV,  xy = TRUE)) +
  # Set aesthetics.
  ggplot2::aes(x, y, alpha = HARV_DSMhill) +
  # Raster plot.
  ggplot2::geom_raster() +
  # Set transparency `alpha` range.
  ggplot2::scale_alpha(range = c(0.15, 0.65),
                       guide = "none")
```

Add crop layer. Here we embed data and aesthetics in each layer as needed.

```{r}
ggplot2::ggplot() +
  
  # Raster crop layer.
  ggplot2::geom_raster(
    data = as.data.frame(DSM_HARV,
                  xy = TRUE),
    ggplot2::aes(x, y, fill = HARV_dsmCrop)) +
  # Color code for crop.
  ggplot2::scale_fill_viridis_c(guide = "none") +
  
  # Raster hill layer.
  ggplot2::geom_raster(
    data = as.data.frame(DSM_hill_HARV,
                  xy = TRUE),
    ggplot2::aes(x, y, alpha = HARV_DSMhill)) +
  # Set transparency `alpha` range.
  ggplot2::scale_alpha(range = c(0.15, 0.65),
                       guide = "none") +

  # Titles and Labels  
  ggplot2::ggtitle("DSM Crop and Hill Plot") +
  ggplot2::labs(x = "UTM Easting Coordinates (m)",
                y = "UTM Northing Coordinates (m)")
```

## Raster Projections

Import DTM data.

```{r}
DTM_HARV <- terra::rast("data/raster/HARV_dtmCrop.tif")
DTM_hill_HARV <- terra::rast("data/raster/HARV_DTMhill_WGS84.tif")
```

These are in different units.

```{r}
terra::crs(DTM_HARV, proj = TRUE)
```

```{r}
terra::crs(DTM_hill_HARV, proj = TRUE)
```


```{r}
ggplot2::ggplot() +
  
  # Raster crop layer.
  ggplot2::geom_raster(
    data = as.data.frame(DTM_HARV,
                  xy = TRUE),
    ggplot2::aes(x, y, fill = HARV_dtmCrop)) +
  # Color code for crop.
  ggplot2::scale_fill_viridis_c(guide = "none") +

  # Titles and Labels  
  ggplot2::ggtitle("DTM Crop Plot, UTM Projection") +
  ggplot2::labs(x = "UTM Easting Coordinates (m)",
                y = "UTM Northing Coordinates (m)")
```

```{r}
ggplot2::ggplot() +

  # Raster hill layer.
  ggplot2::geom_raster(
    data = as.data.frame(
      DTM_hill_HARV,
      xy = TRUE),
    ggplot2::aes(x, y, alpha = HARV_DTMhill_WGS84)) +
  
  # Set transparency `alpha` range.
  ggplot2::scale_alpha(
    range = c(0.15, 0.65),
    guide = "none") +
  ggplot2::scale_fill_gradientn(
    name = "Elevation",
    colors = grDevices::terrain.colors(10)) +

  # Titles and Labels  
  ggplot2::ggtitle("DTM Hill Plot, no Projection") +
  ggplot2::labs(x = "Latitude",
                y = "Longitude")
```

Need to project the hillshade data. Add the projection to the hill data.
Using the UTM projection 18N, `UTMZ18N` that was used for other dataset.

```{r}
terra::crs(
  DTM_hill_HARV,
  proj = TRUE)
```

```{r}
terra::crs(
  terra::project(
    DTM_hill_HARV,
    terra::crs(DTM_HARV)),
  proj = TRUE)
```

```{r}
head(as.data.frame(
      terra::project(
        DTM_hill_HARV,
        terra::crs(DTM_HARV)),
      xy = TRUE))
```

The projection extents are pretty close for these now.

```{r}
terra::ext(DTM_HARV)
```

```{r}
terra::ext(
  terra::project(
    DTM_hill_HARV,
    terra::crs(DTM_HARV)))
```

Now we can define the resolution as 1m.

```{r}
head(as.data.frame(
      terra::project(
        DTM_hill_HARV,
        terra::crs(DTM_HARV),
        res = terra::res(DTM_HARV)),
      xy = TRUE))
```

```{r}
terra::res(DTM_HARV)
```

```{r}
terra::res(
  terra::project(
    DTM_hill_HARV,
    terra::crs(DTM_HARV),
    res = terra::res(DTM_HARV)))
```

```{r}
ggplot2::ggplot() +
  
  # Raster crop layer.
  ggplot2::geom_raster(
    data = as.data.frame(
      DTM_HARV,
      xy = TRUE),
    ggplot2::aes(x, y, fill = HARV_dtmCrop)) +

  # Raster hill layer.
  ggplot2::geom_raster(
    data = as.data.frame(
      # Project Hill data using projection from DTM at resolution 1m.
      terra::project(
        DTM_hill_HARV,
        terra::crs(DTM_HARV),
        res = terra::res(DTM_HARV)),
      xy = TRUE),
    ggplot2::aes(x, y, alpha = HARV_DTMhill_WGS84)) +
  
  # Set transparency `alpha` range.
  ggplot2::scale_alpha(
    name = "Transparency",
    range = c(0.15, 0.65)) +
  # Set gradient with terrain colors.
  ggplot2::scale_fill_gradientn(
    name = "Elevation",
    colors = grDevices::terrain.colors(10)) +

  # Force as rectangle
  ggplot2::coord_sf() +
  # Titles and Labels  
  ggplot2::ggtitle("DTM Crop and Hill Plot") +
  ggplot2::labs(x = "UTM Easting Coordinates (m)",
                y = "UTM Northing Coordinates (m)")
```

## Raster Calculations and Raster Maps

<center>
```
surface - terrain = height

DSM - DTM = CHM (Canopy Height Model)
```
</center>

Examine DTM and DSM data using `terra::crs` and `terra::res` to make sure
they are compatible. Good also to look at them with `plot` or some such.

There are two ways to do raster maps. Start with map operations. 

```{r}
CHM_HARV <- DSM_HARV - DTM_HARV
```

```{r}
ggplot2::ggplot() +
  
  # Raster crop layer.
  ggplot2::geom_raster(
    data = as.data.frame(
      CHM_HARV,
      xy = TRUE),
    ggplot2::aes(x, y, fill = HARV_dsmCrop)) +
  
  # Set gradient with terrain colors.
  ggplot2::scale_fill_gradientn(
    name = "Elevation",
    colors = grDevices::terrain.colors(10)) +

  # Force as rectangle
  ggplot2::coord_sf() +
  # Titles and Labels  
  ggplot2::ggtitle("CHM Crop Plot") +
  ggplot2::labs(x = "UTM Easting Coordinates (m)",
                y = "UTM Northing Coordinates (m)")

```

```{r}
ggplot2::ggplot() +
  ggplot2::geom_density(
    data = as.data.frame(
      CHM_HARV,
      xy = TRUE),
    ggplot2::aes(HARV_dsmCrop))
```

```{r}
ggplot2::ggplot() +
  
  # Raster crop layer.
  ggplot2::geom_raster(
    # Cut values into bins based on `custom_bins` defined above.
    data = 
      dplyr::mutate(
        as.data.frame(CHM_HARV, xy = TRUE),
        canopy_discrete = cut(HARV_dsmCrop, breaks = 4)),
    ggplot2::aes(x, y, fill = canopy_discrete)) +
  
  # Set gradient with terrain colors.
  ggplot2::scale_fill_manual(
    name = "Crop Height",
    values = grDevices::terrain.colors(4)) +

  # Force as rectangle
  ggplot2::coord_sf() +
  # Titles and Labels  
  ggplot2::ggtitle("CHM Crop Plot") +
  ggplot2::labs(x = "UTM Easting Coordinates (m)",
                y = "UTM Northing Coordinates (m)")


```

Another way: `terra::lapp` for more complex operations.
Here we just do the difference of two raster maps, but more complicated 
calculations are possible.

```{r}
CHM_HARV <- terra::lapp(
  terra::sds(
    list(DSM_HARV, DTM_HARV)),
  fun = function(r1, r2) {
    r1 - r2
    })
```

## Export a geotiff

See for instance <https://en.wikipedia.org/wiki/GeoTIFF>.
NB: `*.tif` is old MIME; current preferred file name suffix is `*.tiff`.

```{r eval = FALSE}
terra::writeRaster(CHM_HARV, "CHM_HARV.tiff",
                   filetype = "GTiff",
                   overwrite = TRUE, NAflag = 999)
```

## Multi-band rasters

RGB color photos

Start with one layer. Use `crs`, `res`, etc.

```{r}
terra::rast("data/raster/HARV_RGB_Ortho.tif", lyrs = 1)
```

Now all layers
```{r}
(RGB_stack_HARV <- terra::rast("data/raster/HARV_RGB_Ortho.tif"))
```

```{r}
terra::crs(RGB_stack_HARV, proj = TRUE)
```

```{r}
str(as.data.frame(RGB_stack_HARV, xy = TRUE))
```

Here is a hexplot of two layers. This requires package `hexbin`.
There are nearly 1M points, so helpful to condense in some way.

```{r}
ggplot2::ggplot(
  as.data.frame(
    RGB_stack_HARV,
    xy = TRUE)) +
  ggplot2::aes(HARV_RGB_Ortho_1, HARV_RGB_Ortho_2) +
  ggplot2::geom_hex()
```

While we can plot each layer separately using `geom_raster`.
Instead we use `terra::plotRGB`. Note that this uses the raster data directly
rather than converting first to data frame.

```{r}
terra::plotRGB(RGB_stack_HARV, r = 1, g = 2, b = 3)
```

## Image Stretching

Spread out RGB to use full range of 0-255 for each layer.
Add `stretch` argument to `plotRGB`.

```{r}
terra::plotRGB(RGB_stack_HARV, r = 1, g = 2, b = 3, stretch = "lin")
```

```{r}
terra::plotRGB(RGB_stack_HARV, r = 1, g = 2, b = 3, stretch = "hist")
```

# Day 3 Vector Data

[Lessons 6-9](https://uw-madison-datascience.github.io/r-raster-vector-geospatial/)

[Shape files](https://mapscaping.com/what-is-a-shp-file/) (`*.shp`) require `*.shx` and `*.dbf`, and useful `*.prj` (projections)


```{r}
aoi_boundary_HARV <- sf::st_read(("data/vector/HarClip_UTMZ18.shp"))
```
Examine metadata. 

The number of features in the object (also can use `nrow` and `ncol`).

```{r}
str(aoi_boundary_HARV)
```

```{r}
dim(aoi_boundary_HARV)
```

The type of vector data stored in the object.

```{r}
sf::st_geometry_type(aoi_boundary_HARV)
```

The spatial extent (geographic area covered by) of theobject.

```{r}
sf::st_bbox(aoi_boundary_HARV)
```

The CRS (spatial projection) of the data.

```{r}
sf::st_crs(aoi_boundary_HARV)
```

```{r}
ggplot2::ggplot() +
  # Boundary box.
  ggplot2::geom_sf(
    data = aoi_boundary_HARV,
    linewidth = 1,
    color = "black", fill = "cyan1") +
  ggplot2::ggtitle("AOI Boundary Plot") +
  ggplot2::coord_sf()
```

Read in two more layers

```{r}
lines_HARV <- sf::st_read(("data/vector/HARV_roads.shp"))
```

```{r}
names(lines_HARV)
```

```{r}
footpath_HARV <- subset(lines_HARV, TYPE == "footpath")
```

```{r}
ggplot2::ggplot() +

  # Lines for footpaths
  ggplot2::geom_sf(
    data = footpath_HARV,
    linewidth = 1,
    color = "purple", fill = "blue") +

  # Title and Labels.
  ggplot2::ggtitle("Footpaths") +
  ggplot2::labs(x = "Latitude",
                y = "Longitude") +
  ggplot2::coord_sf()
```


```{r}
ggplot2::ggplot() +

  # Lines for footpaths
  ggplot2::geom_sf(
    data = footpath_HARV,
    ggplot2::aes(color = factor(ResVehic_1)),
    linewidth = 1) +

  # Title and Labels
  ggplot2::ggtitle("Footpaths labeled by ResVehic_1") +
  ggplot2::labs(x = "Latitude",
                y = "Longitude",
                color = "FootpathID") +
  ggplot2::coord_sf()
```

```{r}
points_HARV <- sf::st_read(("data/vector/HARVtower_UTM18N.shp"))
```

Overlay 

```{r}
ggplot2::ggplot() +
  
  # Boundary box.
  ggplot2::geom_sf(
    data = aoi_boundary_HARV,
    linewidth = 1,
    color = "black", fill = "lightgray") +
  
  # Lines for roads.
  ggplot2::geom_sf(
    data = footpath_HARV,
    ggplot2::aes(color = factor(ResVehic_1)),
    linewidth = 1) +
  
  # Points for towers.
  ggplot2::geom_sf(
    data = points_HARV,
    linewidth = 3,
    color = "brown", fill = "red") +
  
  # 
  ggplot2::ggtitle("Footpath, Tower and AOI Boundary") +
  ggplot2::labs(x = "Latitude",
                y = "Longitude",
                color = "Footpath ID") +
  ggplot2::coord_sf()
```

Now look at all roads.

```{r}
linewidth = c(1,1,1.5,2,3)
ggplot2::ggplot() +
  
  # Boundary box.
  ggplot2::geom_sf(
    data = aoi_boundary_HARV,
    linewidth = 1,
    color = "black", fill = "cyan1") +
  
  # Lines for roads.
  ggplot2::geom_sf(
    data = lines_HARV,
    # Color and linewidth by TYPE. The `linewidth` does not apper to work.
    ggplot2::aes(color = factor(TYPE), linewidth = factor(TYPE)))+
  ggplot2::scale_linewidth_manual(values = linewidth, name = "Road Width") +
  
  # Points for towers.
  ggplot2::geom_sf(
    data = points_HARV,
    linewidth = 3,
    color = "brown", fill = "red") +
  
  # 
  ggplot2::ggtitle("Roads, Tower and AOI Boundary") +
  ggplot2::labs(x = "Latitude",
                y = "Longitude",
                color = "Road Type")
```

## Crop Image using Limits

Ideally, would like to be able to change x and y limits, but cannot seem to 
get that to work. Looks like <https://datascience.blog.wzb.eu/2019/04/30/zooming-in-on-maps-with-sf-and-ggplot2/> might have useful insight. Key is that we need to use the same projection.

```{r}
lines_HARV_crop <- sf::st_crop(lines_HARV, 
              xmin = -73, xmax = -72.10,
              ymin = 42.535, ymax = 42.54)
```

```{r}
p <- ggplot2::ggplot() +

  # Lines for footpaths
  ggplot2::geom_sf(
    data = lines_HARV,
    ggplot2::aes(color = factor(TYPE)),
    linewidth = 1)  +

  # Title and Labels
  ggplot2::ggtitle("Footpaths labeled by ResVehic_1") +
  ggplot2::labs(x = "Latitude",
                y = "Longitude",
                color = "FootpathID")
p
```

Crop using `xlim` and `ylim`. Note use of `crs` to get right projection.

```{r}
target_crs <- '+proj=longlat'
p +
  ggplot2::coord_sf(xlim = c(-72.18, -72.17), ylim = c(42.53, 42.54),
                    crs = target_crs)
```

To make this reproducible, would want to have this pick up `crs` from dataset
and figure out croping in reasonable way, say based on subset of roads, or
raster map.

## Overlay Raster and Vector Data

```{r}
p <- ggplot2::ggplot() +
  
  # Raster crop layer.
  ggplot2::geom_raster(
    data = as.data.frame(
      CHM_HARV,
      xy = TRUE),
    ggplot2::aes(x, y, fill = HARV_dsmCrop)) +
  
  # Set gradient with terrain colors.
  ggplot2::scale_fill_gradientn(
    name = "Elevation",
    colors = grDevices::terrain.colors(10)) +

  # Lines for footpaths
  ggplot2::geom_sf(
    data = lines_HARV,
    ggplot2::aes(color = factor(TYPE)),
    linewidth = 1)  +

  # Titles and Labels  
  ggplot2::ggtitle("CHM with Road Overlay") +
  ggplot2::labs(x = "UTM Easting Coordinates (m)",
                y = "UTM Northing Coordinates (m)")
p
```

This should crop the above, but somehow not getting raster plot.

```{r}
target_crs <- '+proj=longlat'
p +
  ggplot2::coord_sf(xlim = c(-72.182, -72.168), ylim = c(42.53, 42.54),
                    crs = target_crs)
```

## US Map

```{r}
state_boundary_US <- 
  sf::st_zm(
    sf::st_read("data/vector/US-State-Boundaries-Census-2014.shp"))
```

```{r}
plot(state_boundary_US)
```

```{r}
ggplot2::ggplot() +
  ggplot2::geom_sf(data = state_boundary_US) +
  ggplot2::ggtitle("Map of US") + 
  ggplot2::coord_sf()
```

```{r}
country_boundary <- 
  # Drop Z dimension.
  sf::st_zm(
    sf::st_read("data/vector/US-Boundary-Dissolved-States.shp"))
```


```{r}
ggplot2::ggplot() +
  ggplot2::geom_sf(data = state_boundary_US) +
  ggplot2::geom_sf(data = country_boundary, fill = "transparent", color = "blue") +
  ggplot2::ggtitle("Map of US") + 
  ggplot2::coord_sf()
```

Note that projection for HARV has different units.

```{r}
sf::st_crs(state_boundary_US)$proj4string
```

```{r}
sf::st_crs(country_boundary)$proj4string
```
```{r}
sf::st_crs(points_HARV)$proj4string
```

Sloppy method that works for mapping on the fly.
However, analytics operations would need to rectify the projections.
Not doing that here.

```{r}
ggplot2::ggplot() +
  ggplot2::geom_sf(data = state_boundary_US) +
  ggplot2::geom_sf(data = country_boundary, fill = "transparent", color = "blue") +
  ggplot2::geom_sf(data = points_HARV, color = "purple") +
  ggplot2::ggtitle("Map of US with HARV site") + 
  ggplot2::coord_sf()
```

